---
title: "Day 9: Statistical Learning — In-Class Activity"
subtitle: "When does a better fit make a worse model?"
author: "Econ 122"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r packages, include=FALSE, eval=TRUE}
mypacks <- c("ggplot2","dplyr","tidyr","ISLR2")
packs <- installed.packages()
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")
lapply(mypacks, library, character.only=TRUE)
```

## Introduction

In class, a very flexible curve fit 20 cars almost perfectly and then did worse
than a straight line on cars it hadn't seen. Now you'll see the same thing with
real wage data, using `Wage` from ISLR2: 3,000 workers' ages and annual wages
(in thousands of dollars).

Imagine you only have a **small survey of 100 workers** to build a model with,
and you want it to predict wages for the **other 2,900**. That's the normal
situation in practice: you fit on the data you have, and the model gets used on
people it has never seen.

**Estimated time: 40 minutes.** The goal for today is **Parts 1–3 and the
wrap-up**. There's a **Bonus** section at the end for anyone who finishes early.
It's optional, so don't worry if you don't get to it.

Each task ends with a question in *italics*. Write a sentence or two; those are
what we'll discuss.

```{r load}
data(Wage)
```

---

## Part 1: Training Error vs. Test Error (12 minutes)

### Task 1.1: Split the data

Run this chunk. It picks 100 random workers for training and keeps everyone else
as a test set.

**What's a seed?** R's "random" numbers come from a fixed recipe that starts at a
number called the *seed*. `set.seed(313)` sets that starting number, so `sample()`
picks the same 100 workers every time you run it, on any computer. Without it,
you'd get a different split on each run. **Don't change the seed** — we want
everyone's numbers to match.

```{r task1_1}
set.seed(313)
train_rows <- sample(nrow(Wage), 100)
wage_train <- Wage[train_rows, ]
wage_test  <- Wage[-train_rows, ]

nrow(wage_train)
nrow(wage_test)
```

### Task 1.2: Write an MSE function

The mean squared error is

$$\text{MSE} = \frac{1}{n}\sum_{i=1}^{n} \big(y_i - \hat y_i\big)^2$$

You'll write it as a **function**: a reusable piece of code that takes inputs
and returns an output. Here's one that averages the gaps between two vectors:

```{r function_example}
average_gap <- function(x, y) {
  gaps <- x - y     # one gap per element
  mean(gaps)        # the last line is what the function returns
}
```

The pieces:

* `average_gap` is the name we give it. `<- function(...)` creates it.
* `x` and `y` are the **arguments**, placeholders for whatever inputs you pass in.
* The code between `{ }` is the **body**. It can have several lines; the value
  of the last line is returned. `gaps` exists only inside the function.

Running that chunk only *defines* the function. To use it, **call** it with
inputs:

```{r function_calls}
average_gap(c(5, 7, 9), c(1, 2, 3))   # gaps are 4, 5, 6
average_gap(c(10, 10), c(4, 8))       # gaps are 6, 2
average_gap(y = c(1, 2, 3), x = c(5, 7, 9))   # name the arguments, any order
```

Notice that `x - y` works element by element: R subtracts the first values, then
the second, and so on. You'll need exactly that for MSE.

Now write `mse(actual, predicted)`. Then run the two checks below (remove the `#`
first). They should return **0** and **1**.

```{r task1_2}
# Your code here


# mse(c(1, 2, 3), c(1, 2, 3))   # should be 0
# mse(c(1, 2, 3), c(2, 3, 4))   # should be 1
```

### Task 1.3: Score one model two ways

Now let's put `mse()` to work on real data. To use it, we need to prepare its
two inputs: the **actual** wages and a model's **predicted** wages for the same
workers.

**Actual wages** are already in the data sets you made in Task 1.1.

**Predicted wages** need a model first. The first line of the chunk below fits
a straight line through the training workers (a linear regression, which we'll
cover on Day 10):

* `lm()` stands for *linear model*.
* `wage ~ age` means "predict `wage` from `age`".
* `data = wage_train` fits it using only the 100 training workers.

The fitted model is saved as `fit1`. Then `predict(fit1, newdata = wage_train)`
gives its predicted wage for every row of `wage_train`.

Your job: compute the MSE on the **training** data and on the **test** data.
For the test MSE, use the same model `fit1`, but swap in `wage_test` in
**both** places.

Fill in the blanks (`___`), then remove the `#`s:

```{r task1_3}
fit1 <- lm(wage ~ age, data = wage_train)

# train_mse_1 <- mse(___, predict(fit1, newdata = ___))
# test_mse_1  <- mse(___, predict(fit1, newdata = ___))
# train_mse_1
# test_mse_1
```

*Which MSE is lower? Why would you expect that, before even running the code?*

---

## Part 2: Turn Up the Flexibility (10 minutes)

Wages tend to rise with age and then level off, so a straight line is too
simple. As in class, a polynomial of degree `d` — `lm(wage ~ poly(age, d))` — can
bend more as `d` grows.

### Task 2.1: Degrees 1 through 10

This is Task 1.3 repeated for degrees 1 through 10, using a `for` loop:

* First we make an empty table, `errors`, with one row per degree (`NA` means
  "not filled in yet").
* `for (d in 1:10) { ... }` runs the code inside `{ }` ten times: once with
  `d = 1`, then `d = 2`, and so on.
* Each time, it fits a degree-`d` model and stores its two MSEs in row `d` of the
  table.

Run it (it uses your `mse()` function, so Task 1.2 has to work first).

```{r task2_1, error = TRUE}
errors <- tibble(degree = 1:10, training = NA, test = NA)

for (d in 1:10) {
  fit <- lm(wage ~ poly(age, d), data = wage_train)
  errors$training[d] <- mse(wage_train$wage, predict(fit, newdata = wage_train))
  errors$test[d]     <- mse(wage_test$wage,  predict(fit, newdata = wage_test))
}

errors
```

### Task 2.2: Plot it

Make a line plot with `degree` on the x-axis, MSE on the y-axis, and one line
each for training and test.

**Hint:** right now training and test MSE sit in two separate columns. To get
one line per set, you need them in a single MSE column, plus a column that says
whether each row is training or test. Use `pivot_longer()` first, then map that
new column to color.

```{r task2_2}
# Your code here

```

*Does training MSE ever go up as the degree rises? Which degree has the lowest
test MSE? Describe what's happening to the degree-10 model in one sentence,
using the word "overfitting."*

---

## Part 3: Make Bias and Variance Yourself (13 minutes)

Part 2 showed **that** test error turns back up. Part 3 shows **why**.

Bias and variance are about what happens **across many possible training sets**.
With real data we only get one. So we'll simulate a world where we know the true
$f$ and can draw as many training sets as we like.

### Task 3.1: The simulated world

Run this chunk. The truth is $f(x) = \sin(2\pi x)$. Each call to
`draw_training_set()` gives you a fresh set of `n` noisy observations from it.

```{r task3_1}
f_true <- function(x) sin(2 * pi * x)

draw_training_set <- function(n = 20) {
  x <- seq(0, 1, length.out = n)
  tibble(x = x, y = f_true(x) + rnorm(n, mean = 0, sd = 0.4))
}

one_set <- draw_training_set()
ggplot(one_set, aes(x, y)) +
  geom_point() +
  geom_function(fun = f_true, linetype = "dashed")
```

Run it two or three more times. There's no seed, so each run draws a new
training set from the same world.

### Task 3.2: Five hundred training sets

Now imagine fitting a degree-1 and a degree-10 polynomial to a training set and
predicting at `x = 0.25`, where the truth is `f_true(0.25)` = 1. The code below
does that for 500 different training sets per model and stores every prediction.
Run it.

```{r task3_2}
predict_at_quarter <- function(degree, n = 20) {
  train <- draw_training_set(n)
  fit <- lm(y ~ poly(x, degree), data = train)
  predict(fit, newdata = tibble(x = 0.25))
}

set.seed(122)
darts <- tibble(
  degree = rep(c(1, 10), each = 500),
  pred   = c(replicate(500, predict_at_quarter(1)),
             replicate(500, predict_at_quarter(10)))
)
```

*Optional, if you're curious how that code works:*

* `predict_at_quarter(degree)` does one round: draw a fresh training set, fit
  the model, and return its prediction at `x = 0.25`.
* `replicate(500, predict_at_quarter(1))` runs that line 500 times and collects
  the 500 results into a vector. Each run draws a new training set, so each
  prediction is different. For example, `replicate(3, rnorm(1))` gives three
  different random numbers.
* `rep(c(1, 10), each = 500)` repeats each value 500 times: five hundred `1`s,
  then five hundred `10`s. That labels each prediction with the degree that
  made it. For example, `rep(c("a", "b"), each = 2)` gives `"a" "a" "b" "b"`.
* `c(...)` stacks the two sets of 500 predictions into one column of 1,000, in
  the same order as the labels.

Now compute, for each degree:

* `bias` = the mean prediction minus the truth, `f_true(0.25)`
* `variance` = the variance of the predictions, `var(pred)`

Then plot a histogram of `pred` for each degree (`facet_wrap(~ degree)`), with a
vertical line at the truth (`geom_vline(xintercept = 1)`).

Fill in the blanks (`___`) and remove the `#`s, then add the histogram:

```{r task3_2_yours}
# bv_20 <- darts %>%
#   group_by(degree) %>%
#   summarise(bias     = ___,
#             variance = ___)
# bv_20

# Your histogram code here

```

**How to read it:** each histogram is the 500 predictions one model made. If
its center sits away from the vertical line at the truth, the model is wrong on
average:
that's **bias**. If it is spread out, the model's answer depends a lot on which
training set it got: that's **variance**.

*Which model has high bias (consistently off in the same direction)? Which has
high variance (jumps around from one training set to the next)? Point to where
you see each one in the histograms.*

### Task 3.3: More data, again

Repeat Task 3.2 with training sets of `n = 200` instead of 20. Fill in both
blanks, remove the `#`s, then compute bias and variance the same way as in
Task 3.2.

```{r task3_3}
set.seed(122)
# darts_200 <- tibble(
#   degree = rep(c(1, 10), each = 500),
#   pred   = c(replicate(500, predict_at_quarter(1,  n = ___)),
#              replicate(500, predict_at_quarter(10, n = ___)))
# )

# Bias and variance: your code here

```

*Which of the two numbers — bias or variance — did more data shrink? Which one
barely moved?*

---

## Wrap-up & Discussion (5 minutes)

1. A classmate says: *"The degree-10 model had the lowest training MSE, so it's
   the least biased model and the best one."* What's right and what's wrong
   about that?

2. Your boss hands you data on **200** customers and asks for the most accurate
   possible spending prediction. A colleague suggests the most flexible method
   available. Using today's results, what would you tell them — and what would
   you want to check before trusting any model?

---

## Bonus (optional)

Only if you've finished everything above. These are extra practice, not part
of today's goal.

### Bonus 1: More wage data

Back to the `Wage` data: what if the training set had **1,500 workers** instead
of 100? This is the code from Tasks 1.1 and 2.1 with new names, so your
100-worker results stay put. Fill in the blank and remove the `#`s.

```{r bonus1}
set.seed(313)
# train_rows_big <- sample(nrow(Wage), ___)
# wage_train_big <- Wage[train_rows_big, ]
# wage_test_big  <- Wage[-train_rows_big, ]
#
# errors_big <- tibble(degree = 1:10, training = NA, test = NA)
# for (d in 1:10) {
#   fit <- lm(wage ~ poly(age, d), data = wage_train_big)
#   errors_big$training[d] <- mse(wage_train_big$wage, predict(fit, newdata = wage_train_big))
#   errors_big$test[d]     <- mse(wage_test_big$wage,  predict(fit, newdata = wage_test_big))
# }
#
# errors       # 100 training workers
# errors_big   # 1,500 training workers
```

*How much worse is the degree-10 model than the best model now, compared to
Part 2? Which did more data shrink here, bias or variance? (Task 3.3 is a hint.)*

### Bonus 2: One split, one answer?

*In Part 2, your "best degree" came from **one** random split of the data. Why
might a different seed give a different answer?* (We'll fix this on Day 14.)
